Validation of GMEA

### 2023-09-14 figs/step_01c_splithcy.Rmd
working on split/analyse

---
title: "GOmeth test"
author: "hcy_meth meta-analysis group"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

@@

Here we will run test enrichment analysis approaches for undderstanding the effect of elevated
homocysteine on blood methylomes.
Homocysteine is thought to be an inhibitor of methyltransferases and hyperhomocysteine is
associated with overall reduced genomic methylation.
Previously we analysed the B-PROOF study data which enabled us to identify probes whose methylation correlatesor anti-correlated with homocysteine level.

The two enrichment analysis approaches we will use are:

* `gsameth`: an ORA approach that addressed the biases inherent with infinium array analysis.

* `gmea`: an experimental application of functional class scoring for infinium analysis.

@@

dm <- read.table("dma3a.tsv")
rownames(dm) <- dm$Row.names
dm$Row.names=NULL

head(dm, 10) %>% kbl() %>% kable_paper("hover", full_width = F)

@@

## ORA with GSAMETH - parth of the MissMethyl package

Now perform ORA analysis using gsameth with the top 500 probes.

```{r,gsameth1}

sigup <- rownames(head(subset(dm,logFC>0),1000))
sigdn <- rownames(head(subset(dm,logFC<0),1000))

tic()
gsaup <- gsameth(sig.cpg=head(sigup,500), all.cpg=rownames(dm), collection=gs_entrez)
gsadn <- gsameth(sig.cpg=head(sigdn,500), all.cpg=rownames(dm), collection=gs_entrez)
toc()

gsaup <- gsaup[order(gsaup$P.DE),]
head(gsaup,10) %>% kbl(caption="Top hypermethylated pathways") %>% kable_paper("hover", full_width = F)

gsadn <- gsadn[order(gsadn$P.DE),]
head(gsadn,10) %>% kbl(caption="Top hypomethylated pathways") %>% kable_paper("hover", full_width = F)

@@

Aggregate probe t-scores to gene level scores.

The object generated has two parts:

1. `df`: a simple table of the gene name and the median tvalue

2. `gme_res_df`: results of 1-sample Wilcox test.

```{r,agg_wt}

CORES= detectCores()

agg <- function(dm) {
  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
  gnl <- mclapply(gnl,unique,mc.cores=CORES)
  dm$UCSC_RefGene_Name <- gnl
  l <- mclapply(1:nrow(dm), function(i) {
    a <- dm[i,]
    len <- length(a[[1]][[1]])
    tvals <- as.numeric(rep(a[2],len))
    genes <- a[[1]][[1]]
    data.frame(genes,tvals)
  },mc.cores=CORES)
  df <- do.call(rbind,l)
  gme_res <- mclapply( 1:length(gn), function(i) {
    g <- gn[i]
    tstats <- df[which(df$genes==g),"tvals"]
    myn <- length(tstats)
    mymean <- mean(tstats)
    mymedian <- median(tstats)
    wtselfcont <- wilcox.test(tstats)
    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
      "p-value(sc)"=wtselfcont$p.value)
  } , mc.cores=CORES )
  gme_res_df <- do.call(rbind, gme_res)
  rownames(gme_res_df) <- gme_res_df[,1]
  gme_res_df <- gme_res_df[,-1]
  tmp <- apply(gme_res_df,2,as.numeric)
  rownames(tmp) <- rownames(gme_res_df)
  gme_res_df <- as.data.frame(tmp)
  gme_res_df$sig <- -log10(gme_res_df[,4])
  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
  gme_res_df$`fdr(sc)` <- p.adjust(gme_res_df$`p-value(sc)`)
  out <- list("df"=df,"gme_res_df"=gme_res_df)
  return(out)
}

tic()
dmagg <- agg(dm)
toc()
head(dmagg$gme_res_df) %>% kbl(caption="top gmea genes (Wilcox)") %>% kable_paper("hover", full_width = F)

@@

This results shows that wilcox test executes faster but 1-sample t-test gives more significant findings.

Check that the distribution is even.

```{r,tmeddist}

@@

If the distribution is even, then established enrichment analysis can be performed.
Options include FGSEA and mitch.

Here I will try downstream analysis with mitch.

```{r,mitch1}

m <- dmagg2$gme_res_df

m <- m[,"median",drop=FALSE]

tic()
mres <- mitch_calc(x=m, genesets=gs_symbols, priority="effect")
toc()

head(mres$enrichment_result,20) %>% kbl(caption="top mitch pathways") %>% kable_paper("hover", full_width = F)
hist(mres$enrichment_result$s.dist) ; grid()

tbl <- mres$enrichment_result

sig <- subset(tbl,p.adjustANOVA<0.05)
nrow(sig)

plot(tbl$s.dist,-log10(tbl$pANOVA),pch=19,col="gray")
grid()
points(sig$s.dist,-log10(sig$pANOVA),pch=19,col="red")

msig <- sig

Some interesting pathways found.

There was no overlap with `gsameth()`.

Interestingly many gene sets exhibit increased methylation.

There could be a directional bias, so I will also conduct a 1-sample t-test.

>>>>

---
title: "Validation of GMEA"
author: "hcy_meth meta-analysis group"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

@@

Previously we demonstrated the ability of GMEA to detect differentially methylated pathways.
There remains a question about how reliable these observations are.
To resolve it, we will use the Hcy dataset which consists of 172 samples.
We will split the data into two groups and analyse these separately.
If the pathway results are similar, then we can be confident about the accuracy.

@@

design <- readRDS("dm3_design.Rds")
dim(design)

dm <- readRDS("dm3_mx.Rds")
dim(dm)

Split the data.

```{r,split1}

g1 <- which(as.data.frame(design)$groups==1)
g2 <- which(as.data.frame(design)$groups==2)
g3 <- which(as.data.frame(design)$groups==3)

seed=100

set.seed(seed); g1s <- sample(g1,floor(length(g1)/2))
g1b <- setdiff(g1,g1s)

set.seed(seed); g2s <- sample(g2,floor(length(g2)/2))
g2b <- setdiff(g2,g2s)

set.seed(seed); g3s <- sample(g3,floor(length(g3)/2))
g3b <- setdiff(g3,g3s)

gxs <- c(g1s,g2s,g3s)
gxs <- gxs[order(gxs)]

gxb <- c(g1b,g2b,g3b)
gxb <- gxb[order(gxb)]

dess <- design[c(gxs),]
dms <- dm[,c(gxs)]

desb <- design[c(gxb),]
dmb <- dm[,c(gxb)]

@@

@@

@@

@@

Conduct a 1-sample t-test.
